### 2023-07-06 paper/localpsf.tex
wordsmithed 5x reduction stuff

	We demonstrate the effectiveness of the proposed method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and inversion for the initial condition in a diffusive transport problem.
	While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. This makes conventional matrix-free Hessian approximation methods expensive in terms of the required number of Hessian operator applications (hence PDE solves). In contrast, the performance of the proposed method depends on the locality and shapes of the impulse responses rather than the numerical rank. Hessian impulse responses typically become more local as the numerical rank increases, which makes high rank Hessians perfect candidates for the proposed approach. 
	We observe that using the proposed PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly 5X--10X reductions in the number of PDE solves required to solve the inverse problems, as compared to classical regularization-based preconditioning and no preconditioning. Finally, we present a comprehensive numerical study for the influence of the diffusion parameters and simulation time (for the diffusive transport problem) on the effectiveness of the approximation. These parameters have an effect on the shape of the impulse responses and on the rank of the Hessian. The results reveal that PSF-based preconditioners are able to form good approximations of high-rank Hessians using a small number of operator applications.
\end{abstract}

@@

Computing the impulse response batches also requires choosing sample point batches via the greedy ellipsoid packing algorithm described in Section~\ref{sec:sample_point_selection}. Choosing the $i$th batch of sample points may require performing up to $\fedim |\pointbatch_i|$ ellipsoid intersection tests, where $|\pointbatch_i|$ is the number of sample points in the $i$th batch. Choosing all of the sample points therefore requires performing at most
%
%
\begin{equation*}
	\fedim \nsamplepts \qquad \text{ellipsoid intersection tests,}
\end{equation*}
%
%
where $\nsamplepts$ is the total number of sample points in all batches. The multiplicative dependence of $\fedim$ with $\nsamplepts$ is undesirable since $\nsamplepts$ may be large, and reducing this cost is possible with more involved computational geometry methods. 
However, from a practical perspective, the cost of choosing sample points is small compared to other parts of the approximation algorithm, and hence such improvements are not pursued here.

@@

In Table~\ref{tab:adv_convergence} we show the number of PCG iterations, $j$, required to solve the Newton linear system to a relative error tolerance of $\|\widehat{\mathbf{q}} - \widehat{\mathbf{q}}_j\| < 10^{-6} \|\widehat{\mathbf{q}}\|$. The solution of the Newton system to which we compare, $\widehat{\mathbf{q}}$, is found via another PCG iteration with a relative residual tolerance of $10^{-11}$. We show results for $T$ ranging from $0.5$ to $2.0$ and $\kappa$ ranging from $10^{-4}$ to $10^{-3}$ using the PSF-based preconditioners with $1$, $5$, and $25$ batches, regularization preconditioning, and no preconditioning. The results show that PSF-based preconditioning outperforms regularization preconditioning and no preconditioning. Adding more batches yields better results, and the impact of adding more batches is more pronounced here than in the ice sheet example. For example, in the mid-range values $T=1.0$ and $\kappa=3.2e{-4}$, PSF(1), PSF(5), and PSF(25) require roughly 1.3X, 2.3X, 5.1X fewer PCG iterations, respectively, as compared to no preconditioning (and greater improvements as compared to regularization preconditioning). The PSF preconditioners perform best in the high rank regime where $T$ is small, which makes sense given that short simulation times yield more localized impulse responses. For example, for $\kappa=3.2e{-4}$, and $T=0.5$, $T=1.0$, and $T=2.0$, using the PSF(5) preconditioner yields 140, 202, and 379 iterations, respectively. The performance of the PSF preconditioners as a function of $\kappa$ is mixed. Reducing $\kappa$ makes the impulse responses thinner and hence easier to fit in batches, but also increases the complexity of the impulse response shapes, which may reduce the accuracy of the RBF interpolation.

@@

We presented an efficient matrix-free PSF method for approximating operators with locally supported non-negative integral kernels. The method requires access to the operator only via application of the operator to a small number of vectors. The idea of the method is to compute batches of impulse responses by applying the operator to Dirac combs of scattered point sources, then interpolate these impulse responses to approximate entries of the operator's integral kernel. The interpolation is based on a new principle we call ``local mean displacement invariance,'' which generalizes classical local translation invariance. The ability to quickly approximate arbitrary integral kernel entries permits us to form an H-matrix approximation of the operator. Fast H-matrix arithmetic is then used to perform further linear algebra operations that cannot be performed easily with the original operator, such as matrix factorization and inversion. The supports of the impulse responses are estimated to be contained in ellipsoids, which are determined a-priori via a moment method that involves applying the operator to a small number of polynomial functions. Point source locations for the impulse response batches are chosen using a greedy ellipsoid packing procedure, in which we choose as many impulse responses per batch as possible, while ensuring that the corresponding ellipsoids do not overlap. We applied the method to approximate the Gauss-Newton Hessians in an ice sheet flow inverse problem governed by the Stokes PDE, and a diffusive transport inverse problem governed by an advection-diffusion PDE. We saw that preconditioners based on the proposed approximation cluster the eigenvalues of the preconditioned Hessian near one, and reduce the required number of PDE solves by factors of roughly 5X--10X. Although the method we presented is not applicable to all Hessians, it is applicable to many Hessians of practical interest. For these Hessians, the proposed method offers a \emph{data scalable} alternative to conventional low rank approximation, due to the ability to form high rank approximations of an operator using a small number of operator applications, and thus forward and adjoint PDE solves.

>>>>

	We demonstrate the effectiveness of the proposed method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and inversion for the initial condition in a diffusive transport problem.
	While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. This makes conventional matrix-free Hessian approximation methods expensive in terms of the required number of Hessian operator applications (hence PDE solves). In contrast, the performance of the proposed method depends on the locality and shapes of the impulse responses rather than the numerical rank. Hessian impulse responses typically become more local as the numerical rank increases, which makes high rank Hessians perfect candidates for the proposed approach. 
	We observe that using the proposed PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly 5X reductions in the number of PDE solves required to solve the inverse problems, as compared to classical regularization-based preconditioning and no preconditioning. Finally, we present a comprehensive numerical study for the influence of the diffusion parameters and simulation time (for the diffusive transport problem) on the effectiveness of the approximation. These parameters have an effect on the shape of the impulse responses and on the rank of the Hessian. The results reveal that PSF-based preconditioners are able to form good approximations of high-rank Hessians using a small number of operator applications.
\end{abstract}

@@

Computing the impulse response batches also requires choosing sample point batches via the greedy ellipsoid packing algorithm described in Section~\ref{sec:sample_point_selection}. Choosing the $i$th batch of sample points may require performing up to $\fedim |\pointbatch_i|$ ellipsoid intersection tests, where $|\pointbatch_i|$ is the number of sample points in the $i$th batch. Choosing all of the sample points therefore requires performing at most
%
%
\begin{equation*}
	\fedim \nsamplepts \qquad \text{ellipsoid intersection tests,}
\end{equation*}
%
%
where $\nsamplepts$ is the total number of sample points in all batches. The multiplicative dependence of $\fedim$ with $\nsamplepts$ is undesirable since $\nsamplepts$ may be large, and reducing this cost is possible with more involved computational geometry methods. 
However, from a practical perspective, the cost of choosing sample points is small compared to other parts of the algorithm, and hence such improvements are not pursued here.

@@

In Table~\ref{tab:adv_convergence} we show the number of PCG iterations, $j$, required to solve the Newton linear system to a relative error tolerance of $\|\widehat{\mathbf{q}} - \widehat{\mathbf{q}}_j\| < 10^{-6} \|\widehat{\mathbf{q}}\|$. The solution of the Newton system to which we compare, $\widehat{\mathbf{q}}$, is found via another PCG iteration with a relative residual tolerance of $10^{-11}$. We show results for $T$ ranging from $0.5$ to $2.0$ and $\kappa$ ranging from $10^{-4}$ to $10^{-3}$ using the PSF-based preconditioners with $1$, $5$, and $25$ batches, regularization preconditioning, and no preconditioning. The results show that PSF-based preconditioning outperforms regularization preconditioning and no preconditioning in all cases except one. The exception is $T=2.0$ and $\kappa=1.0e{-3}$, in which PSF(1) performs slightly worse than regularization preconditioning but better than no preconditioning. Adding more batches yields better results, and the impact of adding more batches is more pronounced here than in the ice sheet example. For example, in the mid-range values $T=1.0$ and $\kappa=3.2e{-4}$, PSF(1), PSF(5), and PSF(25) require 1.3X, 2.3X, 5.1X fewer PCG iterations, respectively, as compared to no preconditioning, and greater improvements as compared to regularization preconditioning. The PSF preconditioners perform best in the high rank regime where $T$ is small, which makes sense given that short simulation times yield more localized impulse responses. For example, for $\kappa=3.2e{-4}$ using the PSF(5) preconditioner yields 140, 202, and 379 iterations for $T=0.5$, $1.0$, and $2.0$, respectively. The performance of the PSF preconditioners as a function of $\kappa$ does not have as clear of a trend. Reducing $\kappa$ makes the impulse responses thinner and hence easier to fit in batches, but also increases the complexity of the impulse response shapes, which may reduce the accuracy of the RBF interpolation. The greatest improvements are seen for $T=0.5$ and $\kappa=1e{-3}$, for which PSF(25) requires roughly 10X fewer PCG iterations than no preconditioning, and roughly 20X fewer iterations than regularization preconditioning.

@@

We presented an efficient matrix-free PSF method for approximating operators with locally supported non-negative integral kernels. The method requires access to the operator only via application of the operator to a small number of vectors. The idea of the method is to compute batches of impulse responses by applying the operator to Dirac combs of scattered point sources, then interpolate these impulse responses to approximate entries of the operator's integral kernel. The interpolation is based on a new principle we call ``local mean displacement invariance,'' which generalizes classical local translation invariance. The ability to quickly approximate arbitrary integral kernel entries permits us to form an H-matrix approximation of the operator. Fast H-matrix arithmetic is then used to perform further linear algebra operations that cannot be performed easily with the original operator, such as matrix factorization and inversion. The supports of the impulse responses are estimated to be contained in ellipsoids, which are determined a-priori via a moment method that involves applying the operator to a small number of polynomial functions. Point source locations for the impulse response batches are chosen using a greedy ellipsoid packing procedure, in which we choose as many impulse responses per batch as possible, while ensuring that the corresponding ellipsoids do not overlap. We applied the method to approximate the Gauss-Newton Hessians in an ice sheet flow inverse problem governed by the Stokes PDE, and a diffusive transport inverse problem governed by an advection-diffusion PDE. We saw that preconditioners based on the proposed approximation cluster the eigenvalues of the preconditioned Hessian near one, and reduce the required number of PDE solves by a factor of roughly 5X. Although the method we presented is not applicable to all Hessians, it is applicable to many Hessians of practical interest. For these Hessians, the proposed method offers a \emph{data scalable} alternative to conventional low rank approximation, due to the ability to form high rank approximations of an operator using a small number of operator applications, and thus forward and adjoint PDE solves.
